---
title: "Assignment_MP2"
author: "Ritik Kumar"
date: "2024-10-25"
output: html_document
---

## Forecast the Insurance Dataset
##### Importing the dataset
```{r}
library(fpp)
library(fpp2)
library(ggplot2)
library(forecast)
library(TTR)
library(readr)

Insurance <- read_csv("C:/Users/malho/Downloads/insurance.csv")
View(Insurance)
```

## 1. Create Timeseries Plot and Inference
##### Considering only Quotes column from January 2002 to April 2005
```{r}
insurance_ts <- ts(Insurance$Quotes, start =c(2002,1), end = c(2005, 4), frequency = 12)

plot(insurance_ts, main = " Monthly Insurance quotations of a US Insurance Company", xlab = "Year", ylab = "Quotes")
```

* we start with plotting the time series for visualize and understand the data.

##### Initial Oberversation 
* We'll assume the data starts from January 2002 and goes monthly
* The data shows lot of fluctuations and variations.
* The plot shows that there is an increasing and decreasing trend in Insurance Quotations from January 2002 to March 2004
* From March 2004 till April 2005, there has been a steady increasing trend in the Insurance quotations.
* from Jan 2002 to April 2005, this data should be good enough to be considered for forecasting.
* This column likely represents the number of insurance quotes or inquiries received.

## 2. Verify how much history to include in your analysis.

```{r}
summary(insurance_ts)
plot(insurance_ts, main = "Monthly Insurance quotations of a US Insurance Company")
```

* The summary function above gives the min, max, mean, median, 1st and 3rd Quarterly values of the times series.
* After that we plot the graph to verify how much history to include in our analysis

## 3. Hypothesize if the dataset has trend, seasonality, or both. 

* This will we check after applying Acf and decomposition for finding trend or seasonality in the dataset

##### 4. Verifying using ACF plot to check for seasonality or autocorrelation
```{r}
Acf(insurance_ts, main =" Acf Plot of Insurance Quotation")
```

##### 5. Verify using decomposition
##### Decompose the time series to verify trend and seasonality
```{r}
decomp <- decompose(insurance_ts)
plot(decomp)
decomp$type
```

* The decomposition seems to be additive. 
* Trend: Long-term movement in the data
* Seasonal: Repeated cycles, usually yearly
* Residual: Random fluctuations that remain after removing trend and seasonality
* Because, with as trend increases, we see there is not a strong increase in the seasonality. The seasonality appears to be repeated cyclically.
* Quotes time series display only weak seasonality.

## 6. Chose an accuracy measure
* We'll use Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) for model evaluation. later

## 7. Create a forecast model for the next 12 months. Include Naive, Average, Exponential Smoothing, HoltWinters, and Decomposition (both types). 

## Naive model
```{r}
naive_forecast <- naive(insurance_ts, h=12)
plot(naive_forecast)
```

##### Naive Method Summary
* The naive model simply assumes the future will be exactly like the most recent value (no trend or seasonality).

## Simple Moving Averages
```{r}
ma5_forecast <- ma(insurance_ts, order =5)
ma9_forecast <- ma(insurance_ts, order =9)
plot(insurance_ts, main ="Plot along with moving averages")
lines(ma5_forecast, col ="Blue")
lines(ma9_forecast, col ="Red")
```

##### Simple Moving Averages Observations
* The average model assumes the future will be the same as the historical average.
* From the plots, it is observed that the higher the order we consider, the smoother the moving average curve in the plot.
* It can be seen that the red line above is the smoothest compared to blue line.
* The Red line (order 9) gives the most real data compared to the other line.  * The higher order averages smoother the plot and do not give the actual values.

## Exponential Smoothing Model (Simple Exponential Smoothing - SES)
```{r}
ses_forecast <- ses(insurance_ts, h=12)
plot(ses_forecast)
attributes(ses_forecast)
```

##### Exponential Smoothing Model Obervations
* Exponential Smoothing gives more weight to recent observations, making it more responsive to changes.
* From Jan 2002 to March 2004 there is observed to be fluctuated trend.After that there is an increasing trend in the data till April 2005. There is a small gap where no trend observed.So, this forecast may not be a right way to forecast.

## Holt-Winters model - Captures both trend and seasonality.
```{r}
hw_model <- hw(insurance_ts)
hw_forecast <- forecast(hw_model, h=12)
plot(hw_forecast, main=" Holt-Winters Forecast")
attributes(hw_forecast)
hw_model$model
```

* Here, The Holt-Winters model considers both trend and seasonality, making it suitable for seasonal data.
* This is because the seasonality isn't increasing with trend. This is an additive time series.

##### Holt-Winters model Observations
* Alpha = 0.9999 Alpha specifies the coefficient for the level smoothing in Holt winters.
* Beta = 0.0272 Beta specifies the coefficient for the trend smoothing in Holt winters. 
* Gamma = 1e-04 Gamma specifies the coefficient for the seasonal smoothing in Holt winters.
* Values 1.0 means that the latest value has highest weight.
* Initial states:
    l = 13.4243 
    b = -0.1468  
    s = 0.236 0.0967 -1.2543 -0.5618 -0.6536 -0.1118
           0.6199 0.8343 -1.2545 -0.4778 1.6549 0.8719
* Sigma = 1.9646 Sigma defines the variance of the forecast values.

## Decomposition-based forecast Model
```{r}
decomp_trend <- tslm(insurance_ts ~ trend + season)
decomp_forecast <- forecast(decomp_trend, h=12)
plot(decomp_forecast, main="Decomposition-Based Forecast")
```

## Decomposition (Additive)
##### Perform additive decomposition Model
```{r}
add_decomp <- decompose(insurance_ts, type= "additive")
plot(add_decomp)
```

* Decomposition splits the time series into trend, seasonal, random and observed components.

## Decomposition (Multiplicative)
##### Perform multiplicative decomposition MOdel
```{r}
mul_decomp <- decompose(insurance_ts, type ="multiplicative")
plot(mul_decomp)
```

* Multiplicative decomposition is used when the seasonal component changes proportionally to the trend.

## 8.Show model rank with accuracy measures (MAE and RMSE)
```{r}
naive_acc <- accuracy(naive_forecast)
naive_acc
ses_acc <- accuracy(ses_forecast)
ses_acc
hw_acc <- accuracy(hw_forecast)
hw_acc
decomp_acc <- accuracy(decomp_forecast)
decomp_acc
```

# Create a summary of accuracy measures (MAE and RMSE)
```{r}
accuracy_summary <- data.frame(
  Model = c("Naive","Exponential Smoothing", "Holt-Winters", "Decomposition"),
  MAE = c(naive_acc[1, "MAE"], ses_acc[1, "MAE"], hw_acc[1, "MAE"], decomp_acc[1, "MAE"]),
  RMSE = c(naive_acc[1, "RMSE"], ses_acc[1, "RMSE"], hw_acc[1, "RMSE"], decomp_acc[1, "RMSE"])
)

print(accuracy_summary)
```

* This table ranks the models based on their accuracy using MAE and RMSE. Lower values indicate better accuracy.

## 9. Choose which models and how are you going to use them for Forecasting
* Based on the accuracy measures, we would select the model with the lowest MAE and RMSE, typically the Holt-Winters or Exponential Smoothing model.

##### Best & Worst Forecasts

* To start with, there is nothing like best or worst forecast.
* Considering the accuracy data above, HoltWinters forecast seems to fit the time series the best as it has the least error values.
* And decomposition forecast seems to be the worst as it has the largest ME and RMSE values.

## 10. Provide the forecast for the next 12 months (point and range) and explain why you feel confident with these forecasts
```{r}
final_forecast <- forecast(hw_model, h=12)
print(final_forecast)
plot(final_forecast, main="Final Forecast for the Next 12 Months")
```

* This is the final forecast, providing point estimates for the next 12 months along with confidence intervals.

## Conclusion 
* The data seemed to have strong trend and but weak seasonality initially and we checked the same with Acf and confirmed it.
* Based on the forecasting models like naive, Exponential Smoothing Model, and Holt Winters, we can see that Holt-winters forecast provides to be the better forecasting method in this case.
* This is because the forecast fits perfectly and also the error values are quiet low for Holt Winters forecast.
* This shows that our hypothesis is correct based on the accuracy of all the models.
* Based on the analysis and forecast, the time series will same steady over the next year and the next 2 years.